Ecological network control

### 2018-06-03 paper/reviewers-response.Rmd

---
title: "Ecological network control"
subtitle: "JEcol-2017-0720 - Detailed response to comments"
author: "Fernando Cagua, Kate L. Wootton, Daniel B. Stouffer"
classoption: a4paper
csl: journal-of-ecology.csl
documentclass: artikel1
output: 
  bookdown::pdf_document2: 
    keep_tex: yes
    toc: false
    number_sections: false
header-includes:
  - \usepackage{setspace}
  - \usepackage{xr}
  - \externaldocument[M-]{manuscript}
  - \externaldocument[S-]{supp-info}
  - \usepackage[most]{tcolorbox}
  - \definecolor{block-gray}{gray}{0.9}
  - \newtcolorbox{myquote}{colback=block-gray,grow to right by=-10mm,grow to left by=-10mm, boxrule=0pt,boxsep=0pt,breakable}
---

\doublespacing

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, 
                      warning = FALSE)
```

We are thankful to the editors, David Gibson and Nicole Rafferty, and two anonymous reviewers for the insightful comments to improve the manuscript. 
In the following sections, you can find a detailed response to each comment raised during the previous round of review.

@@

# Editor comments

> "I agree with the AE and reviewers that this is an interesting and potentially important manuscript. I support the AE's decision to reject but invite resubmission for all the reasons mentioned. I encourage resubmission if you feel that you can better bring out the conceptual importance of their work for plant ecologists. If you wish to strengthen the applied focus with specific conservation recommendations then you may wish to submit to Journal of Applied Ecology. Reviewer 1 offers suggestions for both approaches. I hope that you choose the former."

Thanks to the Senior Editor, David Gibson, for his comments and his willingness to allow resubmission. 
After lengthy consideration of the the two recommended alternatives, we have chosen to focus on the conceptual importance of our work in line with the editor's stated preference. 
To do this effectively, we ultimately ended up rewriting both the manuscript and the supplementary materials almost completely. We think the extra effort was worth it as we feel the current version does a much better job at explaining the utility and insights generated by the proposed approach.

#  Associate Editor comments

@@

> "Both reviewers see potential value in the application of control theory to species interaction networks and found the work to be of interest. I agree that the method and its application to plant-pollinator networks are interesting, particularly the findings that contrast invaded vs uninvaded networks and use novel metrics to describe the influence of invasive species in networks. Both reviewers also note, however, that the current ms falls short of clearly and convincingly demonstrating the utility of control theory in making management decisions. They call for more in-depth exploration and illustration of what novel and important contributions this method would make to conservation. For these reasons, this work may be better-suited to a more applied journal (e.g., Journal of Applied Ecology), assuming the ms is revised in the recommended direction."

We thank the Associate Editor, Nicole Rafferty, for the constructive feedback and highlighting the strengths of the paper. 
As we noted above in our response to the Senior Editor, we opted to use our revisions to strengthen the conceptual component rather than emphasising its application in specific management and conservation scenarios. 
We opted to take this decision for two main reasons. 
First, the empirical data we use does not explicitly include any management or conservation scenarios. 
As a consequence and despite the potential of the approach, we currently lack *direct* evidence of the utility of the approach for management decisions.
Second, we think that highlighting the conceptual approach offered a greater opportunity to draw attention to the strongest/most interesting aspects of the paper. 
Namely, the contrasts between invaded and uninvaded communities and the role of invasive species, as the reviewer points out.

XXX At end of last sentence above, which reviewer?

@@

> "The management of complex ecological networks is of vast importance for the conservation of the natural world. While the field of ecological networks is vast few have tackled this question in a practical way. This paper would make a significant contribution to a field that needs to change direction if it is to help inform conservation as often purported in studies from this field."

We are grateful to the reviewer for the positive feedback and encouraging words.

> "The control theory approach is very interesting. I think however there needs to be more clarity in the paper. It is hard to read. A reworking of it including using descriptions of the variable throughout instead of variable names would improve this."

Previously we used two sections to explain our approach: a section called "Theoretical framework" where we described the control methods, and another called "Methods" where we described their application on empirical data. 
At that moment, we thought this was the best approach; in hindsight, we realise that separating these two aspects made it hard to understand the core control concepts and their relationship with ecology. 
As mentioned by the reviewer, this problem was compounded by abundant jargon and mathematical notation. 
To improve clarity, we have now combined the two sections into a single "Methods" section in which we explain the theory alongside with the empirical application. 
We also spend more time providing intuitive explanations of the concepts of control theory and have moved the most  technical details to the supplementary information. This latter move (shifting the technical details to the supplementary information) allowed us to use far less jargon and simplify the notation. We hope that the reviewer agrees that the new version is easier to follow.

@@

> "Better context needs to be given for previous work to look at the management of networks in ecology. There are a few papers and knowing how this one differs from these (not just the ones that look at node importance for example) is important. Some of these papers are mentioned in the discussion but in the intro, it is important to know what they do and why this paper is novel."

Thanks for this observation. 
In line with feedback from both editors, we have chosen to emphasise on the conceptual aspects of the work and reduced the core question of management applications to appear only in the discussion. 
Nevertheless, we have also rewritten the introduction and took care to be clear at explaining how this study relates to and differs from previous work.

@@

XXX answer missing XXX

# Reviewer comments #2

> "I read with great interest the manuscript by Cagua et al. on 'Quantifying the manageability of pollination networks in an invasion context'. The authors use control theory to develop an approach to assess how ecological networks – here pollination networks – can be managed, and test the approach on ten pairs of invaded and uninvaded plant-pollinator communities. Overall, the methodological approach appears to be sound (I’m not familiar with control theory, so I’ll have to leave in-depth assessment to an expert), the descriptions and rationale behind the assumptions used to develop the methods are logical and reasonable, and the illustration of the concept using published networks seems appropriate. Also, although the manuscript may require a few minor edits it is well and clearly written. I do, however, have some concerns about the applicability of the approach and the findings, which I’ll describe in more detail below."

We are thankful to the reviewer for indicating the aspects which they believe were strengths of our work. We maintain the methodological approach in the latest version, and have used these details to determine which elements were essential to maintain in the current revision.

### 2018-06-04 paper/manuscript.Rmd

The first step in applying methods of control theory is to construct a directed network that is able to provide an indication of the extent to which species affect each other's abundance. 
In some ecological networks, establishing the directionality can be relatively straightforward, for example when links represent biomass transfer or energy flow [@isbell_human_2013]. 
In pollination networks, however, this directionality is less obvious as both species can, in theory, benefit from the interaction. 
We overcome that obstacle by noting that the extent to which species *i* affects species *j* relative to the extent to which *j* affects *i* can be summarised by the interaction asymmetry [@bascompte_asymmetric_2006]. 
This asymmetry is given by $$a(i,j) = a(j,i) = \frac{\left | d_{ij}-d_{ji} \right |}{max\left ( d_{ij}, d_{ji} \right )},$$ where the dependence of plant *i* on pollinator *j*, *d~ij~*, is the proportion of the visits from pollinator *j* compared to all pollinator visits to plant *i*. 
Previous research has shown that mutualistic interactions are often highly asymmetric in natural communities; in other words, if a plant species is largely dependent on a pollinator species, then that pollinator tends to depend rather weakly on the plant (or vice versa). 
We, therefore, create a directed link between species *i* and *j* when $d_{ij}-d_{ji} \geq 0$ to establish the most likely direction of control between a species pair (Figure \@ref(fig:fig-example-networks)a). 
By basing the direction of the links on the asymmetry of their dependence, we are able to generate a network that is consistent with the dynamics of the community while satisfying the requirements of structural controllability.

@@

We then examined whether any species-level structural properties could predict a species control capacity $\phi$ and the likelihood of being a superior node.
*TODO: Still need to decide how to do this and how to compare traditional centrality metrics to the new control based ones. Will probably base the comparison on stability/feasibility metrics or/and on coextinction of species groups.*Maximum matchings and control configurations. In directed networks, a maximum matching is the largest possible set of links that do not share start or end nodes (dark purple). Maximum matchings are not necessarily unique, instead, each of them is related to a possible minimum driver node set in the network (the nodes to which an external control input, in orange, should be applied in order to ensure controllability). The size of the minimum driver node set *D* corresponds exactly to the number of unmatched nodes (the number of nodes in the network *N* minus the matching size). To account for network size, we use the size of the minimum driver node set relative to the total number of nodes *n~D~ = D/N* as a measure of the extent to which the network structure can be harnessed to control the system.

>>>>

The first step in applying methods of control theory is to construct a directed network that is able to provide an indication of the extent to which species affect each other's abundance. 
In some ecological networks, establishing the directionality can be relatively straightforward, for example when links represent biomass transfer or energy flow [@isbell_human_2013]. 
In pollination networks, however, this directionality is less obvious as both species can, in theory, benefit from the interaction. 
We overcome that obstacle by noting that the extent to which species *i* affects species *j* relative to the extent to which *j* affects *i* can be summarised by the interaction asymmetry [@bascompte_asymmetric_2006]. 
This asymmetry is given by $$a(i,j) = a(j,i) = \frac{\left | d_{ij}-d_{ji} \right |}{max\left ( d_{ij}, d_{ji} \right )},$$ where the dependence of plant *i* on pollinator *j*, *d~ij~*, is the proportion of the visits from pollinator *j* compared to all pollinator visits to plant *i*. 
Previous research has shown that mutualistic interactions are often highly asymmetric in natural communities; in other words, if a plant species is largely dependent on a pollinator species, then that pollinator tends to depend rather weakly on the plant (or vice versa). 
We, therefore, create a directed link between species *i* and *j* when $d_{ij}-d_{ji} \geq 0$ to establish the most likely direction of control between a species pair (Figure \@ref(fig:fig-example-networks)a). 
Sometimes, however, there is no assymetry in the dependences between species pairs ($d_{ij}=d_{ji}$), and we cannot infer the direction of control. 
As such, we deem both species to be equally likely to affect each other and create two directed links between them, one from *i* to *j* and another from *j* to *i*. 
By basing the direction of the links on the asymmetry of their dependence, we are able to generate a network that is consistent with the dynamics of the community while satisfying the requirements of structural controllability.

@@

To examine whether any species-level structural properties could predict a species control capacity $\phi$ and the likelihood of being a superior node $sigma$, we calculated these properties for each of our networks.
In the networks that contained reciprocal links (because there was not assymetry in the dependences of a species pair), we averaged a species' $\phi$ and $sigma$ across every possible "non-reciprocal" realisation of the network (more details can be found in the Supplementary Information section \@ref(S-reciprocal-links)). 
*TODO: Still need to decide how to do this and how to compare traditional centrality metrics to the new control based ones. Will probably base the comparison on stability/feasibility metrics or/and on coextinction of species groups.*

### 2018-06-04 paper/supp-info.Rmd

>>>>

## Networks with reciprocal links {#reciprocal-links}

```{r}
drake::loadd(directed_networks)
recipr <- directed_networks %>%
  purrr::map(igraph::reciprocity, mode = "ratio") %>%
  unlist() %>% {
    dplyr::data_frame(net_name = names(.), prop_bid = .)
  } %>%
  dplyr::inner_join(metadata, by = "net_name") %>%
  dplyr::filter(study %in% c("bartomeus", "lopezaraiza"))
```

Our procedures to find minimum driver node sets is only adequate when there are no reciprocal (bidirectional) links in the network. 
The proportion of reciprocal links in relatively low. 
In fact the proportion of species pairs that are reciprocally connected in our empirical networks is just `r round(mean(recipr$prop_bid)*100, 2)`%. 
However, reciprocal links are not a rare occurrence as `r sum(recipr$prop_bid != 0)` out of `r nrow(recipr)` empirical networks had at least one. 
To calculate our controllability metrics at both the network (the relative size of the minimum driver node set, $n_D$) and species level (control capacity, $\phi$, and the likelihood of being a superior node, $\sigma$) we first need to generate all possible versions of the of the network that include only non-reciprocal links.

Based on the available information, all of these network is as likely to be one that best represent the superior/matched relationships in the network. 
As such, we average the three metrics ($n_D$, $\phi$, and $\sigma$) across all the networks without reciprocal links without weighting by the number of possible control configurations.

```{r}
drake::loadd(fig_bidirectional)
```

```{r fig-bidirectional, fig.height=fig_bidirectional$height, fig.width=fig_bidirectional$width,fig.cap = "In order to calculate the controllability metrics for a network with reciprocal links (left), we first decompose the network into several versions that contain no reciprocal links (right). As each of these networks represent a different equiprobable scenario, the controllability metrics for the network with reciprocal links are the unweighted mean of these metrics across the non-reciprocal networks."}
replayPlot(fig_bidirectional$plot)
```
